The Power of Two Random Choices\\ \large A follow up seminar to the Parallel Algorithms lecture\footnote{Jesper Larsson Tr\"aff and Francesco Versaci, lecture ``Parallel Algorithms'', 2012 winter term at TU Wien.}

### 2014-02-14 article/article.tex
use align* instead of align

\usepackage{color}
\newcommand\todo[1]{\textcolor{red}{(TODO: #1)}}
\newcommand\load{L_{\mathrm{max}}}

@@

\subsection{Balls-into-bins model}
\label{sec:model}
Although several different variants of this simple balls-into-bins game exist, we will assume a very basic model to highlight the crucial ideas in both placement strategies as well as the analysis of them. In our model, we start with $n$ bins and put balls into them, one per time step $t$. We assume that at any time step $t$ at most $m = h n$ balls are in the bins, where $h$ is some constant greater or equal to $1$. At each time step $t$, we delete one randomly chosen ball from the bins. While most of the paper and especially the analysis focuses on the lightly loaded case where $m=n$, we will give results for the case $m > n$ as well. Throughout the paper, we will denote the maximum load by $\load$ and the additive gap by $\delta$. Obviously, 
\begin{align}
\delta = \load - \frac{m}{n}.
\end{align}

\subsection{Single-choice balls-into-bins}
\label{sec:single-choice}
A simple strategy for the balls-into-bins problem is to choose one bin uniformly and independently at random for each ball, which we will call the \emph{single-choice balls-into-bins} strategy. Using this strategy, the maximum load of the fullest bin will be
\begin{comment}
\footnote{A more accurate result exists: The maximum load is $\load = \Gamma^{-1}\left(n\right)\left(1+O\left(\frac{1}{\log \Gamma^{-1}\left(n\right)}\right)\right)$ \cite{G91}.}
\end{comment}
\begin{align}
\load = \frac{\ln n}{\ln \ln n} \cdot \left(1 + o\left(1\right)\right)
\end{align}
for the case $m=n$, with high probability\footnote{We say that an event $\mathcal E$ occurs \emph{with high probability} (w.h.p.) if $\Pr\left[\mathcal E \right] \geq 1 - n^{-\alpha}$ for an arbitrary constant $\alpha > 0$.}~\cite{RS98}. This implies\footnote{For a detailed proof, see Theorem \ref{theorem:asymptotic} in Appendix \ref{sec:asymptotic}.}
\begin{align}
\load\in \Theta\left(\frac{\ln n}{\ln \ln n}\right).
\end{align}
Note that since the average load is 1, $\delta = L_{max}$ in this case. For the heavily loaded case, w.h.p.~the maximum gap is at most~\cite{RS98}.
\begin{align}
\delta = \Theta\left(\sqrt{\frac{m \cdot \log n}{n}}\right).
\end{align}

\subsection{Multiple-choice balls-into-bins}
Consider a variation of the above \emph{single-choice balls-into-bins} strategy: For each ball, chose $d \geq 2$ bins independently and uniformly at random and place the ball into the \emph{least full bin}\footnote{If there are multiple bins with the same minimal load, chose one among them arbitrarily.}. This simple \emph{multiple-choice balls-into-bins} scheme, which we will call the \emph{Greedy} algorithm\footnote{A detailed definition will follow in Section \ref{sec:greedy}}, achieves a gap
\begin{align}
\delta = \frac{\log \log n}{\log d} + O(1)
\end{align}
 w.h.p.~for all $m$~\cite{ABKU99}~\cite{BCSV06}.

@@

Consider a slightly more involved strategy, which we will call the \emph{Always-go-left} scheme: Suppose that the $n$ bins are partitioned evenly into $d$ groups, each of size\footnote{For the sake of a clear presentation, we will assume that $d$ divides $n$ evenly in the remainder of this paper.} $\lfloor \frac{n}{d}\rfloor$ or $\left\lceil\frac{n}{d}\right\rceil$. We select one bin from each group uniformly and independently at random and put the ball into the least full of those $d$ bins. In case a tie occurs, put the ball into the \emph{left-most} bin. That is, assuming a fixed ordering of bins $b_i$ for $1 \leq i \leq n$, from among the $d' \leq d$ bins with the same minimal load, we put the ball into the bin with smallest index $i$. Somewhat surprisingly, V\"ocking \cite{VOC03} showed that such a strategy employing \emph{asymmetric tie-breaking} in combination with \emph{non-uniform} bin selection reduces the bound for the gap to\footnote{The symbol $\Phi_d$ denotes a generalization of the golden ratio, based on generalized Fibonacci numbers. Refer to Appendix \ref{sec:fibonacci} for the definition.}
\begin{align}
\delta = \frac{\ln\ln n}{d \cdot \ln \Phi_d} + O(1)
\end{align}
w.h.p. Not only is this a significant reduction of the asymptotic bound, V\"ocking was able to show that the Always-go-left strategy is optimal up to additive constants \cite{VOC03}.

@@

\section{Classes of random placement strategies}
\label{sec:classesOfPlacement}
Let $B$ be the set of bins $\{1,2...n\}$, $\Omega = B^{d}$ be the sample space and $\mathcal{F} = 2^{\Omega}$ be the set of measurable events. We are only interested in events that represent a set of bins of size $d$. Thus, let $\mathcal{F}_d = \{\mathcal{E} \in \mathcal{F}: \left\vert \mathcal{E} \right\vert = d\}$ and set $\Pr\left(\mathcal{E}  \right) = 0$  $\forall \mathcal{E} \in \mathcal{F} \setminus \mathcal{F}_d $. The three classes of algorithms are distinguished by the probability function $\Pr\left(\mathcal{E}\right)$ for events $\mathcal{E} \in \mathcal{F}_d$:
\begin{compactitem}
\item  \emph{Uniform and independent.} Choose each of the $d$ bins for a ball uniformly and independently at random from the set of bins $B$. This implies 
\begin{align}
\forall \mathcal{E} \in \mathcal{F}_d: \Pr\left(\mathcal{E}\right) = \frac{1}{n ^{d}}
\end{align}
\item Class 2: \emph{Non-uniform and independent.} Choose the i-th bin from $B$ independently at random according to some probability function
\begin{align}
\Pr: B \rightarrow \left[0,1\right]
\end{align}

\item Class 3: \emph{Non-uniform and dependent.} Choose the $d$ bins for a ball from $\Omega$ at random:
\begin{align}
\Pr: B^{d} \rightarrow \left[0,1\right]
\end{align}
\end{compactitem}

@@

\begin{theorem}
\label{theorem:greedy}
Consider any sequence of insertions and deletions s.t.~at most $hn$ balls are in the bins at any time. If the balls are placed into bins by the Greedy algorithm, the maximum load is at most 
\begin{align}
\frac{m}{n} + \frac{\log \log n}{\log d}+ O(h)
\end{align}
w.h.p.~at any time $t$ \cite{ABKU99} \cite{BCSV06}.
\end{theorem}

@@

\begin{theorem}
\label{theorem:agln}
If $n$ balls are placed into $n$ bins using the Always-go-left scheme, the number of balls in the fullest bin will be 
\begin{align}
\frac{\ln\ln n}{d \cdot \ln \Phi_d} + O(1)
\end{align}
w.h.p.
\end{theorem}

\begin{theorem}
\label{theorem:algm}
Consider any sequence of insertions and deletions s.t.~at most $hn$ balls are in the bins at any time. If the balls are placed into bins by the Greedy algorithm, at any time $t$ the maximum load is at most 
\begin{align}
\frac{\ln\ln n}{d \cdot \ln \Phi_d} + O(h)
\end{align}
w.h.p.
\end{theorem}

@@

We will prove the bounds for $m=n$ for both the Greedy and the Always-go-left scheme in Sections \ref{sec:analysis:greedy} and \ref{sec:analysis:alg}, respectively. However, we use a simplifying assumption \ref{assumption:independence} in those two sections to provide a more clear presentation of the proof technique. Section \ref{sec:analysis:nondistinctBalls} will extend these proofs to remove the simplifying assumption we make initially. The witness tree method can be used to show that the same bounds hold for the heavily loaded case where $m> n$ too, but we do not give such a proof in this paper. Instead, we refer the interested reader to V\"ocking \cite{VOC03}. Berenbrink et al.  ~\cite{BCSV06}, Talwar et al.~\cite{TW13} and Mitzenmacher \cite{MRS01} provide proofs the heavily loaded case using different approaches as well.

\subsection{Greedy}
\label{sec:analysis:greedy}
In this Section, we will give a detailed proof for Theorem \ref{theorem:greedy}. However, we will restrict ourselves to the case where $m = n$ and will make use of the symplifying assumption \ref{assumption:independence}.

\subsubsection{Symmetric witness trees}
\label{sec:analysis:definitionSymWT}
A \emph{symmetric witness tree} of order $L$ is a complete $d$-ary tree with $d^{L}$ leaves. Each node $v$ represents a ball $b_v = \mathrm{ball}(v)$ . The same ball may be represented by several nodes. The nodes and edges of the witness tree represent events that may occur or not. For simplicity, we assume that all events represented by a witness tree are stochastically independent.

We define edge and leaf events:
\begin{compactitem}
\item \emph{Edge event:} Consider some edge $e = (u,v)$ of the witness tree, s.t.~$v$ is the i-th child of $u$. Then $e$ represents the event that the i-th location of the ball $b_u = \mathrm{ball}(u)$ points to the same bin as one of the locations of the ball $b_v = \mathrm{ball}(v)$. 
\todo{illustrate}
\item \emph{Leaf event:} A leaf node $v$ of the witness tree represents the event that each of the $d$ locations of the ball $b_v = \mathrm{ball}(v)$ points to a bin that contains at least three additional balls (i.e., balls that are not represented by a node of the tree) at the time of the insertion of the ball $b_v$.
\todo{illustrate}
\end{compactitem}

@@

The definition for leaf events refers to bins containing some number of balls at some specified time. To avoid dependencies, the crucial trick is that edge events are defined in terms of the alternative locations of balls, not by their finally assigned bin.

@@

\subsubsection{Construction}
\label{sec:analysis:constructionSymWT}
We construct a witness tree for the event that some bin contains $L+4$ balls as follows:
\begin{compactenum}
\item The topmost ball $b_1$ in bin $x$ is assigned to the root node $v_r$. Note that $b_1$ is the last ball that was inserted into the bin $x$.
\item Since the Greedy-scheme puts ball $b_1$ into bin $x$ only if each of the $d$ sampled locations contain at least the same number of balls at bin $x$, we can conclude that each of the $d$ locations contain at least $L+3$ balls at the time ball $b_1$ was inserted. Thus, the $d$ children of the root node are assigned the topmost balls of these $d$ bins.
\item Consider these child nodes and continue the assignment of the previous step recursively until the leaf nodes are reached.  
\end{compactenum}

By construction, all edge events of the witness tree are activated. Note that the ball assigned to the root node is on top of at least $L+3$ balls. The balls assigned to the root's child node are each on top of $L+2$ or more balls and so on. It follows that each ball assigned to a leaf node is on top of at least 3 other balls, implying that the $d$ locations of each leaf node point to bins with at least 3 balls in each. This implies the activation of all leaf nodes of the witness tree.

\subsubsection{Probability of activation}
\label{sec:analysis:probabilitySymWT}
In this section, we derive the probability that some witness tree of order $L$ is activated, which gives an upper bound for the probability that the maximum load is $L+3$ at some time $t$. The probability $p_a$ that some witness tree of order $L$ is activated is bounded by the number of witness trees $w$ times the upper bound on the probability $p_1$ that a specific witness tree is activated.

Initially, we will make use of the following assumption in the analysis. 
\begin{assumption}
\label{assumption:independence}
All events represented by the nodes and edges of a witness tree are stochastically independent. 
\end{assumption}
In other words, we assume that no ball is referenced by more than one node of a witness tree. Obviously, a correct analysis will have to consider balls assigned to more than one node, but this simplifying assumption is useful to highlight the differences between the symmetric and asymmetric allocation schemes. Section \ref{sec:analysis:nondistinctBalls} will extend the analysis to witness trees with redundant balls.

@@

An edge\footnote{where $v$ is the i-th child of $u$} $(u, v)$ is activated with probability at most $d/n$, which is the probability that the i-th location of the ball $b_u = \mathrm{ball}(u)$ points to a particular location of the ball $b_v = \mathrm{ball}(v)$ times the number of locations $d$. This implies a probability of at most
\begin{align}
p_e = \left(\frac{d}{n}\right)^{k-1}
\end{align} 
or the event that all $k-1$ edges of a witness tree are activated. Note that  the assumption that all balls are distinct (Assumption \ref{assumption:independence}) implies that all edge events are independent.

The probability that a leaf event is activated is bounded above by $3^{-d}$. Recall that by the definition of leaf events, each of the $d$ locations of the ball $b_{vl} = \mathrm{ball}(v_l)$ has to point to a bin with at least 3 additional balls in it. The number of those bins is bounded by $\lfloor n/3 \rfloor$ at all time steps $t$, implying a probability of at most 
\begin{align}
\frac{n/3}{n} = \frac{1}{3}
\end{align}
that one such bin is chosen. This implies a probability of at most $3^{-d}$ that all $d$ bins are chosen among those with at least 3 additional bins in them. Thus, the probability that all leaf events are activated is at most $p_l = 3^{-d\cdot l}$, where $l = d^{L}$ is the number of leaves in a witness tree.

Thus, the probability that an activated symmetric witness tree exists at time $t$ is at most 
\begin{align}
p_{swt} &= w \cdot p_e \cdot p_l \\
		&= n^{k} \cdot \left(\frac{d}{n}\right)^{k-1} \cdot 3^{-d \cdot l} \\
		&= n \cdot d^{k-1} \cdot 3^{-d \cdot l}
\end{align}
Using $k \leq 2l$ we get
\begin{align}
p_{swt} &\leq n \cdot d^{2\cdot l} \cdot 3^{-d \cdot l} \\
		&= n \cdot \left(d^2\right)^l \cdot \left( 3^{-d} \right)^l
\end{align}
which we simplify by using $2d^2 \leq 3^d$
\begin{align}
p_{swt} &\leq n \cdot \left(\frac{3^d}{2}\right)^l \cdot \left( 3^{-d}\right)^l \\
		&= n \cdot 2^{-l} \cdot 3^{d\cdot l} \cdot 3^{-d \cdot l} \\
		&= n\cdot 2^{-l}		\\
		&= n\cdot 2^{-d^L}		
\end{align}

For any constant $\alpha > 0 $, let
\begin{align}
L &\geq \log_d\log_2 n + \log_d\left(1+\alpha\right)
\end{align}
and note that
\begin{align}
L &\geq  \frac{\ln\log_2 n}{\ln d} + \frac{\ln(1+\alpha)}{\ln d} \\
   &= \frac{\ln \log_2 n}{\ln d} \cdot \left(1 + \frac{\ln\left(1+\alpha\right)}{\ln \log_2 n}\right) \\
  &=  \frac{\ln \ln n}{\ln d} \cdot \left( 1+ o\left( 1\right)\right)
\end{align}

Then, 
\begin{align}
p_{swt} &\leq n \cdot 2^{{-d}^{\log_d\log_2 n + \log_d\left(1+\alpha\right)}} \\
		&= n \cdot 2^{-{d}^{\log_d\log_2 n}\cdot d^{\log_d \cdot\left(1+\alpha\right)}} \\
		&= n \cdot \left( 2^{-\log_2 n \cdot \left(1+ \alpha \right)} \right) \\
		&= n \cdot n^{-\left(1+\alpha\right)} = n^{-\alpha}
\end{align}

@@

For each node we define a label $(h, i)$, where $h, i \in \mathbb{N}$, $0 \leq h \leq L$ and  $1\leq i \leq d$. 
\begin{compactitem}
\item A node that is the root of a Fibonacci tree $T_d(1)$ is assigned the label $(0,1)$.
\item Analogously, a node that is the root of a Fibonacci tree $T_d(2)$ is assigned the label $(0,2)$. Thus, nodes labeled $(0,1)$ or $(0,2)$ do not have children.
\item A node with label $(0, i)$, where $i>2$, has $i-1$ children with labels $(0,i-1)\dots (0,1)$
\item A node with label $(h,i)$ , where $h>0$, has $d$ children with the labels $(h,i-1)\dots(h,1)$,$(h-1,d)\dots(h-1, i)$.
\item The root of a Fibonacci tree $T_d(d\cdot h+i)$ has the label $(h, i)$. 
\end{compactitem}
Thus, our the root node of our witness tree will be labeled $(L, 1)$. Note that the labels are unique within the set of children of a node $v$, but not for all nodes of the tree.

\subsubsection{Construction}
\label{sec:analysis:constructionAsymWT}
The labels defined in the previous section are useful to state the following invariant, which will be maintained while assigning balls to the nodes of the witness tree.

@@

\subsubsection{Full witness trees}
\label{sec:analysis:FullWT}
Let $gamma \geq 2$ be a suitable constant. The root node of a \emph{full witness tree} has $\gamma$ children, denoted by $v_{c_i}$ with $1\leq i \leq \gamma$. The nodes $v_{c_i}$ have exactly one child each and the balls assigned to them are pairwise distinct. The root thus has $\gamma$ grandchildren, each of which is the root of a witness tree of order $L$. The time constraints activation of a full witness tree is defined analogously to the previous variants.

>>>>

\usepackage{color}
\newcommand\todo[1]{\textcolor{red}{(TODO: #1)}}
\newcommand\load{L_{\mathrm{max}}}
\newcommand\numberthis{\addtocounter{equation}{1}\tag{\theequation}}
% within an align* environment, number and label the equation. #1 is the label text to be appended to eqn:
\newcommand\neqn[1]{\numberthis\label{eqn:#1}}

@@

\subsection{Balls-into-bins model}
\label{sec:model}
Although several different variants of this simple balls-into-bins game exist, we will assume a very basic model to highlight the crucial ideas in both placement strategies as well as the analysis of them. In our model, we start with $n$ bins and put balls into them, one per time step $t$. We assume that at any time step $t$ at most $m = h n$ balls are in the bins, where $h$ is some constant greater or equal to $1$. At each time step $t$, we delete one randomly chosen ball from the bins. While most of the paper and especially the analysis focuses on the lightly loaded case where $m=n$, we will give results for the case $m > n$ as well. Throughout the paper, we will denote the maximum load by $\load$ and the additive gap by $\delta$. Obviously, 
\begin{align*}
\delta = \load - \frac{m}{n}.
\end{align*}

\subsection{Single-choice balls-into-bins}
\label{sec:single-choice}
A simple strategy for the balls-into-bins problem is to choose one bin uniformly and independently at random for each ball, which we will call the \emph{single-choice balls-into-bins} strategy. Using this strategy, the maximum load of the fullest bin will be
\begin{comment}
\footnote{A more accurate result exists: The maximum load is $\load = \Gamma^{-1}\left(n\right)\left(1+O\left(\frac{1}{\log \Gamma^{-1}\left(n\right)}\right)\right)$ \cite{G91}.}
\end{comment}
\begin{align*}
\load = \frac{\ln n}{\ln \ln n} \cdot \left(1 + o\left(1\right)\right)
\end{align*}
for the case $m=n$, with high probability\footnote{We say that an event $\mathcal E$ occurs \emph{with high probability} (w.h.p.) if $\Pr\left[\mathcal E \right] \geq 1 - n^{-\alpha}$ for an arbitrary constant $\alpha > 0$.}~\cite{RS98}. This implies\footnote{For a detailed proof, see Theorem \ref{theorem:asymptotic} in Appendix \ref{sec:asymptotic}.}
\begin{align*}
\load\in \Theta\left(\frac{\ln n}{\ln \ln n}\right).
\end{align*}
Note that since the average load is 1, $\delta = L_{max}$ in this case. For the heavily loaded case, w.h.p.~the maximum gap is at most~\cite{RS98}.
\begin{align*}
\delta = \Theta\left(\sqrt{\frac{m \cdot \log n}{n}}\right).
\end{align*}

\subsection{Multiple-choice balls-into-bins}
Consider a variation of the above \emph{single-choice balls-into-bins} strategy: For each ball, chose $d \geq 2$ bins independently and uniformly at random and place the ball into the \emph{least full bin}\footnote{If there are multiple bins with the same minimal load, chose one among them arbitrarily.}. This simple \emph{multiple-choice balls-into-bins} scheme, which we will call the \emph{Greedy} algorithm\footnote{A detailed definition will follow in Section \ref{sec:greedy}}, achieves a gap
\begin{align*}
\delta = \frac{\log \log n}{\log d} + O(1)
\end{align*}
 w.h.p.~for all $m$~\cite{ABKU99}~\cite{BCSV06}.

@@

Consider a slightly more involved strategy, which we will call the \emph{Always-go-left} scheme: Suppose that the $n$ bins are partitioned evenly into $d$ groups, each of size\footnote{For the sake of a clear presentation, we will assume that $d$ divides $n$ evenly in the remainder of this paper.} $\lfloor \frac{n}{d}\rfloor$ or $\left\lceil\frac{n}{d}\right\rceil$. We select one bin from each group uniformly and independently at random and put the ball into the least full of those $d$ bins. In case a tie occurs, put the ball into the \emph{left-most} bin. That is, assuming a fixed ordering of bins $b_i$ for $1 \leq i \leq n$, from among the $d' \leq d$ bins with the same minimal load, we put the ball into the bin with smallest index $i$. Somewhat surprisingly, V\"ocking \cite{VOC03} showed that such a strategy employing \emph{asymmetric tie-breaking} in combination with \emph{non-uniform} bin selection reduces the bound for the gap to\footnote{The symbol $\Phi_d$ denotes a generalization of the golden ratio, based on generalized Fibonacci numbers. Refer to Appendix \ref{sec:fibonacci} for the definition.}
\begin{align*}
\delta = \frac{\ln\ln n}{d \cdot \ln \Phi_d} + O(1)
\end{align*}
w.h.p. Not only is this a significant reduction of the asymptotic bound, V\"ocking was able to show that the Always-go-left strategy is optimal up to additive constants \cite{VOC03}.

@@

\section{Classes of random placement strategies}
\label{sec:classesOfPlacement}
Let $B$ be the set of bins $\{1,2...n\}$, $\Omega = B^{d}$ be the sample space and $\mathcal{F} = 2^{\Omega}$ be the set of measurable events. We are only interested in events that represent a set of bins of size $d$. Thus, let $\mathcal{F}_d = \{\mathcal{E} \in \mathcal{F}: \left\vert \mathcal{E} \right\vert = d\}$ and set $\Pr\left(\mathcal{E}  \right) = 0$  $\forall \mathcal{E} \in \mathcal{F} \setminus \mathcal{F}_d $. The three classes of algorithms are distinguished by the probability function $\Pr\left(\mathcal{E}\right)$ for events $\mathcal{E} \in \mathcal{F}_d$:
\begin{compactitem}
\item  \emph{Uniform and independent.} Choose each of the $d$ bins for a ball uniformly and independently at random from the set of bins $B$. This implies 
\begin{align*}
\forall \mathcal{E} \in \mathcal{F}_d: \Pr\left(\mathcal{E}\right) = \frac{1}{n ^{d}}
\end{align*}
\item Class 2: \emph{Non-uniform and independent.} Choose the i-th bin from $B$ independently at random according to some probability function
\begin{align*}
\Pr: B \rightarrow \left[0,1\right]
\end{align*}

\item Class 3: \emph{Non-uniform and dependent.} Choose the $d$ bins for a ball from $\Omega$ at random:
\begin{align*}
\Pr: B^{d} \rightarrow \left[0,1\right]
\end{align*}
\end{compactitem}

@@

\begin{theorem}
\label{theorem:greedy}
Consider any sequence of insertions and deletions s.t.~at most $hn$ balls are in the bins at any time. If the balls are placed into bins by the Greedy algorithm, the maximum load is at most 
\begin{align*}
\frac{m}{n} + \frac{\log \log n}{\log d}+ O(h)
\end{align*}
w.h.p.~at any time $t$ \cite{ABKU99} \cite{BCSV06}.
\end{theorem}

@@

\begin{theorem}
\label{theorem:agln}
If $n$ balls are placed into $n$ bins using the Always-go-left scheme, the number of balls in the fullest bin will be 
\begin{align*}
\frac{\ln\ln n}{d \cdot \ln \Phi_d} + O(1)
\end{align*}
w.h.p.
\end{theorem}

\begin{theorem}
\label{theorem:algm}
Consider any sequence of insertions and deletions s.t.~at most $hn$ balls are in the bins at any time. If the balls are placed into bins by the Greedy algorithm, at any time $t$ the maximum load is at most 
\begin{align*}
\frac{\ln\ln n}{d \cdot \ln \Phi_d} + O(h)
\end{align*}
w.h.p.
\end{theorem}

@@

We will prove the bounds for $m=n$ for both the Greedy and the Always-go-left scheme in Sections \ref{sec:analysis:greedy} and \ref{sec:analysis:alg}, respectively. However, we use a simplifying Assumption \ref{assumption:independence} in those two sections to provide a more clear presentation of the proof technique. Section \ref{sec:analysis:nondistinctBalls} will extend these proofs to remove the simplifying assumption we make initially. The witness tree method can be used to show that the same bounds hold for the heavily loaded case where $m> n$ too, but we do not give such a proof in this paper. Instead, we refer the interested reader to V\"ocking \cite{VOC03}. Berenbrink et al.  ~\cite{BCSV06}, Talwar et al.~\cite{TW13} and Mitzenmacher \cite{MRS01} provide proofs the heavily loaded case using different approaches as well.

\subsection{Greedy}
\label{sec:analysis:greedy}

\subsubsection{Symmetric witness trees}
\label{sec:analysis:definitionSymWT}
A \emph{symmetric witness tree} of order $L$ is a complete $d$-ary tree with $d^{L}$ leaves. Each node $v$ represents a ball $b_v = \mathrm{ball}(v)$ . The nodes and edges of the witness tree represent events that may occur or not. The same ball may be represented by several nodes. However, initially we will make use of the following assumption in the analysis.
\begin{assumption}
\label{assumption:independence}
All events represented by the nodes and edges of a witness tree are stochastically independent. 
\end{assumption}
In other words, we assume that no ball is referenced by more than one node of a witness tree. Obviously, a correct analysis will have to consider balls assigned to more than one node, but this simplifying assumption is useful to highlight the differences between the symmetric and asymmetric allocation schemes. Section \ref{sec:analysis:nondistinctBalls} will extend the analysis to witness trees with redundant balls.

We define edge and leaf events:
\begin{compactitem}
\item \emph{Edge event:} Consider some edge $e = (u,v)$ of the witness tree, s.t.~$v$ is the i-th child of $u$. Then $e$ represents the event that the i-th location of the ball $b_u = \mathrm{ball}(u)$ points to the same bin as one of the locations of the ball $b_v = \mathrm{ball}(v)$. 
\todo{illustrate}
\item \emph{Leaf event:} A leaf node $v$ of the witness tree represents the event that each of the $d$ locations of the ball $b_v = \mathrm{ball}(v)$ points to a bin that contains at least three additional balls (i.e., balls that are not represented by a node of the tree) at the time of the insertion of the ball $b_v$.
\todo{illustrate}
\end{compactitem}
The definition for leaf events refers to bins containing some number of balls at some specified time. To avoid dependencies, the crucial trick is that edge events are defined in terms of the alternative locations of balls, not by their finally assigned bin.

We say that an edge or leaf event is \emph{activated} if the associated event occurs. A witness tree is activated when all its edge and leaf events are activated.

@@

@@

\begin{lemma}
When using the Greedy algorithm to put balls into bins and arbitrary $L > 0$, a bin containing more than $L+3$ balls implies the activation of a symmetric witness tree of order $L$.
\end{lemma}

\begin{proof}
We construct a witness tree for the event that some bin contains $L+4$ balls as follows:
\begin{compactenum}
\item The topmost ball $b_1$ in bin $x$ is assigned to the root node $v_r$. Note that $b_1$ is the last ball that was inserted into the bin $x$.
\item Since the Greedy-scheme puts ball $b_1$ into bin $x$ only if each of the $d$ sampled locations contain at least the same number of balls at bin $x$, we can conclude that each of the $d$ locations contain at least $L+3$ balls at the time ball $b_1$ was inserted. Thus, the $d$ children of the root node are assigned the topmost balls of these $d$ bins.
\item Consider these child nodes and continue the assignment of the previous step recursively until the leaf nodes are reached.  
\end{compactenum}

By construction, all edge events of the witness tree are activated. Note that the ball assigned to the root node is on top of at least $L+3$ balls. The balls assigned to the root's child node are each on top of $L+2$ or more balls and so on. It follows that each ball assigned to a leaf node is on top of at least 3 other balls, implying that the $d$ locations of each leaf node point to bins with at least 3 balls in each. This implies the activation of all leaf nodes of the witness tree.
\end{proof}

\subsubsection{Probability of activation}
\label{sec:analysis:probabilitySymWT}
In this section, we derive the probability that some witness tree of order $L$ is activated, which gives an upper bound for the probability that the maximum load is $L+3$ at some time $t$.

\begin{lemma}\label{lemma:swt:activation}
For some arbitrary $\alpha$ and large enough $L$, the activation of a symmetric witness tree of order $L$ occurs with probability at most $n^{-\alpha}$.
\end{lemma}
Speaking less formaly, Lemma \ref{lemma:swt:activation} states that the for large enough $L$, the activation of a symmetric witness tree is unlikely.

\begin{proof}
The probability $p_{swt}$ that some symmetric witness tree of order $L$ is activated is bounded by the number of witness trees $w$ times the upper bound on the probability $p_1$ that a specific witness tree is activated.

@@

An edge\footnote{where $v$ is the i-th child of $u$} $(u, v)$ is activated with probability at most $d/n$, which is the probability that the i-th location of the ball $b_u = \mathrm{ball}(u)$ points to a particular location of the ball $b_v = \mathrm{ball}(v)$ times the number of locations $d$. This implies a probability of at most
\begin{align*}
p_e = \left(\frac{d}{n}\right)^{k-1}
\end{align*} 
or the event that all $k-1$ edges of a witness tree are activated. Note that  the assumption that all balls are distinct (Assumption \ref{assumption:independence}) implies that all edge events are independent.

The probability that a leaf event is activated is bounded above by $3^{-d}$. Recall that by the definition of leaf events, each of the $d$ locations of the ball $b_{vl} = \mathrm{ball}(v_l)$ has to point to a bin with at least 3 additional balls in it. The number of those bins is bounded by $\lfloor n/3 \rfloor$ at all time steps $t$, implying a probability of at most 
\begin{align*}
\frac{n/3}{n} = \frac{1}{3}
\end{align*}
that one such bin is chosen. This implies a probability of at most $3^{-d}$ that all $d$ bins are chosen among those with at least 3 additional bins in them. Thus, the probability that all leaf events are activated is at most $p_l = 3^{-d\cdot l}$, where $l = d^{L}$ is the number of leaves in a witness tree.

Thus, the probability that an activated symmetric witness tree exists at time $t$ is at most 
\begin{align*}
p_{swt} &= w \cdot p_e \cdot p_l \\
		&= n^{k} \cdot \left(\frac{d}{n}\right)^{k-1} \cdot 3^{-d \cdot l} \\
		&= n \cdot d^{k-1} \cdot 3^{-d \cdot l}
\end{align*}
Using $k \leq 2l$ we get
\begin{align*}
p_{swt} &\leq n \cdot d^{2\cdot l} \cdot 3^{-d \cdot l} \\
		&= n \cdot \left(d^2\right)^l \cdot \left( 3^{-d} \right)^l
\end{align*}
which we simplify by using $2d^2 \leq 3^d$
\begin{align*}
p_{swt} &\leq n \cdot \left(\frac{3^d}{2}\right)^l \cdot \left( 3^{-d}\right)^l \\
		&= n \cdot 2^{-l} \cdot 3^{d\cdot l} \cdot 3^{-d \cdot l} \\
		&= n\cdot 2^{-l}		\\
		&= n\cdot 2^{-d^L}		
\end{align*}

For any constant $\alpha > 0 $, let
\begin{align*}
L &\geq \log_d\log_2 n + \log_d\left(1+\alpha\right) \neqn{swt:chooseL}
\end{align*}

Then, 
\begin{align*}
p_{swt} &\leq n \cdot 2^{{-d}^{\log_d\log_2 n + \log_d\left(1+\alpha\right)}} \\
		&= n \cdot 2^{-{d}^{\log_d\log_2 n}\cdot d^{\log_d \cdot\left(1+\alpha\right)}} \\
		&= n \cdot \left( 2^{-\log_2 n \cdot \left(1+ \alpha \right)} \right) \\
		&= n \cdot n^{-\left(1+\alpha\right)} = n^{-\alpha} \neqn{swt:result}
\end{align*}
Thus, for any constant $\alpha$ and large enough $L$ as specified in Equation \ref{eqn:swt:chooseL} the probability that a symmetric witness tree of order $L$ is activated is bounded above by $p_{swt }\leq n^{-\alpha}$.
\end{proof}

Since the probabilty for the activation of a symmetric witness tree of order $L$ upper-bounds the probability of the bad event that some bin contains more than $L+3$ balls, Equation \ref{eqn:swt:result} also is an upper bound for the latter. Note that Equation \ref{eqn:swt:chooseL} can be written as
\begin{align*}
L &\geq  \frac{\ln\log_2 n}{\ln d} + \frac{\ln(1+\alpha)}{\ln d} \\
   &= \frac{\ln \log_2 n}{\ln d} \cdot \left(1 + \frac{\ln\left(1+\alpha\right)}{\ln \log_2 n}\right) \\
  &=  \frac{\ln \ln n}{\ln d} \cdot \left( 1+ o\left( 1\right)\right) \neqn{swt:boundL}
\end{align*}
This proves Theorem \ref{theorem:greedy} for the case $m=n$, under the assumption that all events are stochastically independent (Assumption \ref{assumption:independence}). \qed

@@

\begin{lemma}
When using the Always-go-left algorithm to put balls into bins and arbitrary $L > 0$, a bin containing more than $L+3$ balls implies the activation of an asymmetric witness tree of order $L$.
\end{lemma}
\begin{proof}
For each node we define a label $(h, i)$, where $h, i \in \mathbb{N}$, $0 \leq h \leq L$ and  $1\leq i \leq d$. 
\begin{compactitem}
\item A node that is the root of a Fibonacci tree $T_d(1)$ is assigned the label $(0,1)$.
\item Analogously, a node that is the root of a Fibonacci tree $T_d(2)$ is assigned the label $(0,2)$. Thus, nodes labeled $(0,1)$ or $(0,2)$ do not have children.
\item A node with label $(0, i)$, where $i>2$, has $i-1$ children with labels $(0,i-1)\dots (0,1)$
\item A node with label $(h,i)$ , where $h>0$, has $d$ children with the labels $(h,i-1)\dots(h,1)$,$(h-1,d)\dots(h-1, i)$.
\item The root of a Fibonacci tree $T_d(d\cdot h+i)$ has the label $(h, i)$. 
\end{compactitem}
Thus, our the root node of our witness tree will be labeled $(L, 1)$. Note that the labels are unique within the set of children of a node $v$, but not for all nodes of the tree. The labels are useful to state the following invariant, which will be maintained while assigning balls to the nodes of the witness tree.

@@

\subsubsection{Full witness trees}
\label{sec:analysis:FullWT}
Let $\gamma \geq 2$ be a suitable constant. The root node of a \emph{full witness tree} has $\gamma$ children, denoted by $v_{c_i}$ with $1\leq i \leq \gamma$. The nodes $v_{c_i}$ have exactly one child each and the balls assigned to them are pairwise distinct. The root thus has $\gamma$ grandchildren, each of which is the root of a witness tree of order $L$. The time constraints and activation of a full witness tree is defined analogously to the previous variants.
